---
title: "Jannink and Wu 2003 MCMC methods"
author: "Frederick J. Boehm"
date: "1/19/2020"
output: html_document
params:
  lastmod: !r lubridate::now()
bibliography: references.bib  
---

Last modified: `r params$lastmod`.

IN the terminology of @jannink2003estimating, the sequence of updates per iteration is:

1. update QTL inheritance matrix $Q$  
1. update QTL allelic effects $a$  
1. update family means $\beta$ and residual variance $\sigma^2$  
1. update configuration given number of alleles, ie, $C | l$  
1. update the number of alleles $l$

The later @jannink2004optimal uses a similar sequence of updates:

1. Update the number of QTL alleles 
1. Update the configuration given the number of alleles
1. update QTL variance (when there are multiple QTL)
1. update allelic effects
1. update position and genotypes jointly
1. update family means









## Simulate traits

```{r load_attie}
geno <- readRDS("../data/derived_data/attie_geno.rds")
```

```{r}
library(magrittr)
```

```{r}
config <- matrix(nrow = 8, ncol = 2)
config[, 1] <- rep(c(0, 1), each = 4)
config[, 2] <- 1 - config[, 1]
# simulate a trait
tr <- reducedscan::sim1(aprobs = geno$`1`[ , , 100], allelic_series = config, allelic_effects = c(-30, 30), error_variance = 1)
s1_out <- qtl2::scan1(genoprobs = geno, pheno = tr)
map <- readRDS("../data/derived_data/attie_map.rds")
plot(s1_out, map = map, chr = 1)
peak_index <- qtl2::find_peaks(s1_out, map = map) %>%
  dplyr::filter(chr == 1) %>%
  dplyr::select(pos) %>%
  unlist() %>%
  (function(x) which(map$`1` == x))
# get the founder allele effects
effects <- qtl2::scan1coef(genoprobs = geno[ , 1], pheno = tr) %>%
  (function(x) qtl2pleio::get_effects(marker_index = peak_index, allele_effects_matrix = x, map = map$`1`))

```

```{r mcmc}
set.seed(2020-02-04)
mcmc_out_poisson2 <- reducedscan::jannink_mcmc(niter = 10000, 
                                      initial_values = list(configuration = diag(8), 
                                                            effects = effects, 
                                                            residual_variance = 1), 
                                      genoprobs = geno$`1`[ , , 100], 
                                      trait = tr, 
                                      allelic_number_prior = "poisson", 
                                      poisson_prior_mean = 2)
mcmc_out_uniform <- reducedscan::jannink_mcmc(niter = 10000, 
                                      initial_values = list(configuration = diag(8), 
                                                            effects = effects, 
                                                            residual_variance = 1), 
                                      genoprobs = geno$`1`[ , , 100], 
                                      trait = tr, 
                                      allelic_number_prior = "uniform", 
                                      poisson_prior_mean = 2)
```


I'm starting to think that I should add an option to use a truncated poisson prior on the number of alleles. It seems that I routinely get a result with exactly four alleles. I think this is related to the fact that the second Stirling number for 8 is maximized at 4, by a large amount.

If I use a poisson prior on the number of alleles, then I'll drive the number of alleles to a smaller value.



