---
title: "Jannink and Wu 2003 MCMC methods"
author: "Frederick J. Boehm"
date: "1/19/2020"
output: html_document
params:
  lastmod: !r lubridate::now()
bibliography: references.bib  
---

Last modified: `r params$lastmod`.

IN the notation of @jannink2003estimating, the sequence of updates per iteration is:

1. update QTL inheritance matrix $Q$  
1. update QTL allelic effects $a$  
1. update family means $\beta$ and residual variance $\sigma^2$  
1. update configuration given number of alleles, ie, $C | l$  
1. update the number of alleles $l$

The later @jannink2004optimal uses a similar sequence of updates:

1. Update the number of QTL alleles 
1. Update the configuration given the number of alleles
1. update QTL variance (when there are multiple QTL)
1. update allelic effects
1. update position and genotypes jointly
1. update family means









## Simulate traits

```{r load_attie}
geno <- readRDS("../data/derived_data/attie_geno.rds")
```

```{r}
library(magrittr)
```

```{r}
config <- matrix(nrow = 8, ncol = 2)
config[, 1] <- rep(c(0, 1), each = 4)
config[, 2] <- 1 - config[, 1]
# simulate a trait
tr <- reducedscan::sim1(aprobs = geno$`1`[ , , 100], allelic_series = config, allelic_effects = c(-10, 10), error_variance = 1)
s1_out <- qtl2::scan1(genoprobs = geno, pheno = tr)
map <- readRDS("../data/derived_data/attie_map.rds")
plot(s1_out, map = map, chr = 1)
peak_index <- qtl2::find_peaks(s1_out, map = map) %>%
  dplyr::filter(chr == 1) %>%
  dplyr::select(pos) %>%
  unlist() %>%
  (function(x) which(map$`1` == x))
# get the founder allele effects
s1c_out <- qtl2::scan1coef(genoprobs = geno[ , 1], pheno = tr)
qtl2::plot_coefCC(s1c_out, map)
effects <- s1c_out %>%
  (function(x) qtl2pleio::get_effects(marker_index = peak_index, allele_effects_matrix = x, map = map$`1`))

```

```{r mcmc}
set.seed(2020-02-04)
mcmc_out_poisson2 <- reducedscan::jannink_mcmc(niter = 10000, 
                                      initial_values = list(configuration = diag(8), 
                                                            effects = effects, 
                                                            residual_variance = 1), 
                                      genoprobs = geno$`1`[ , , 100], 
                                      trait = tr, 
                                      allelic_number_prior = "poisson", 
                                      poisson_prior_mean = 2)
mcmc_out_poisson2_rv <- reducedscan::jannink_mcmc_rv_only(niter = 10000, 
                                      initial_values = list(configuration = config, 
                                                            effects = c(-10, 10), 
                                                            residual_variance = 10), 
                                      genoprobs = geno$`1`[ , , 100], 
                                      trait = tr, 
                                      allelic_number_prior = "poisson", 
                                      poisson_prior_mean = 2)
mcmc_out_poisson2_rv_fx <- reducedscan::jannink_mcmc_rv_fx_only(niter = 10000, 
                                      initial_values = list(configuration = config, 
                                                            effects = c(-10, 10), 
                                                            residual_variance = 3), 
                                      genoprobs = geno$`1`[ , , 100], 
                                      trait = tr, 
                                      allelic_number_prior = "poisson", 
                                      poisson_prior_mean = 2)



mcmc_out_uniform <- reducedscan::jannink_mcmc(niter = 10000, 
                                      initial_values = list(configuration = diag(8), 
                                                            effects = effects, 
                                                            residual_variance = 1), 
                                      genoprobs = geno$`1`[ , , 100], 
                                      trait = tr, 
                                      allelic_number_prior = "uniform", 
                                      poisson_prior_mean = 2)
mcmc_out_poisson3 <- reducedscan::jannink_mcmc(niter = 10000, 
                                      initial_values = list(configuration = diag(8), 
                                                            effects = effects, 
                                                            residual_variance = 1), 
                                      genoprobs = geno$`1`[ , , 100], 
                                      trait = tr, 
                                      allelic_number_prior = "poisson", 
                                      poisson_prior_mean = 3)
```

```{r}


```


## Open questions about the inferences

1. Role of starting values
  1. residual variance
  1. configuration
  1. allelic number
  1. effects
1. Role of true effects
1. Role of poisson prior mean hyperparameter
1. How do mistaken inferences in configuration affect statistical power in QTL scans?
  1. Too many alleles, but otherwise reasonable groupings of founders
  1. Too few alleles
  1. Too many alleles and unreasonable groupings of founders
  

  
